---
title: "Trimming parameter changes | fastq-mcf -s flag"
subtitle: "Linda Crnic Institute for Down Syndrome"
author: "Kohl Kinning"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  html_notebook: 
    code_folding: none
    theme: cosmo
    toc: yes
    toc_float:
      collapsed: yes
editor_options: 
  chunk_output_type: inline
---

The `-s` (scale) parameter was set to `0` in the pipeline. This parameter is interpreted as log2(s), so the setting of `0` results in adapter trimming after a single base match.  We noticed over-aggressive trimming of the ends of reads due to such a low threshold for matching. In this report, we initially look at the differences of several `-s` settings and then continue to look at the differences between a stting of `0` and `1.6`.

### Libraries

```{r}
library(ggplot2)
library(ggpubr)
library(dplyr)
library(plotly)
```

### Difference in total counts 

A match of 1 and 2 bases looks similar, while higher settings result in a sharp dropoff.
```{r}
trim_counts <- read.table("~/Projects/CD4_CD8_NK_trimming_investigations/trimming_results.csv", header=TRUE, sep=",")

ggplot(trim_counts, aes(x=as.character(scale), y=clipped_end_reads)) +
  geom_bar(stat="identity") +
  xlab("-s parameter") + ylab("Clipped reads") +
  ggtitle("Effect of -s parameter on read clipping") +
  theme_pubr()
```

### Read the counts data in

Look at the HTSeq resulting gene count data for the setting of `0` and `1.6`.
```{r}
zero_counts <- read.table("~/Projects/CD4_CD8_NK_trimming_investigations/counts/0/CD4_100B2_HTSeq_counts.txt", row.names=1)
one_six_counts <- read.table("~/Projects/CD4_CD8_NK_trimming_investigations/counts/1.6/CD4_100B2_HTSeq_counts.txt", row.names=1)

bound_counts <- zero_counts
names(bound_counts) <- "lower_thresh"
bound_counts$higher_thresh <- one_six_counts$V2
bound_counts$diff <- bound_counts$lower_thresh - bound_counts$higher_thresh
bound_counts$ratio_change <- bound_counts$diff/bound_counts$lower_thresh
```



### Scatter plot

Observe no outliers when comparing the counts between the settings. Any changes are minor and in a consistent direction.
```{r}
ggplotly(
ggplot(bound_counts, aes(x=log2(lower_thresh), y=log2(higher_thresh), label=rownames(bound_counts))) +
  geom_point(shape=1) +
  xlab("log2(-s 0 resulting counts)") + ylab("log2(-s 1.6 resulting counts)") +
  ggtitle("Difference in resulting counts\ndue to -s parameter") +
  theme_pubr()
)
```




### Total difference in counts

Look at the total difference in HTSeq counts due to the `-s` parameter change
```{r}
sum(bound_counts$lower_thresh)-sum(bound_counts$higher_thresh)
```


### Zero count genes

Number of genes lost completely
```{r}
colSums(bound_counts[1:2]==0)
```

### Average change

Look at the average per-gene counts change, using the absolute differences
```{r}
mean(abs(bound_counts$diff))
mean(abs(bound_counts$ratio_change))
```



### Gene gain

Which genes do we gain with the higher thresh?
```{r}
gained_genes <-
  setdiff(
    rownames(bound_counts[which(bound_counts$lower_thresh==0),]),
    rownames(bound_counts[which(bound_counts$higher_thresh==0),])
  )

bound_counts[gained_genes,1:3]
```



### Gene loss

Which genes do we lose with the higher thresh?
```{r}
lost_genes <-
  setdiff(
    rownames(bound_counts[which(bound_counts$higher_thresh==0),]),
    rownames(bound_counts[which(bound_counts$lower_thresh==0),])
  )

bound_counts[lost_genes,1:3]
```


### Highest diffrences

Look at the genes which show the largest differences in counts
```{r}
bound_counts %>% rownames_to_column() %>% top_n(50, diff) %>% arrange(-diff)
```

